\documentclass[letterpaper, 11 pt]{ximera}
\input{../lessonplanheader.tex}
\input{../printstyles.tex}


\title{Acknowledgements}


\begin{document}


These notes served as course notes for the Spring 2024 Number Theory course at Davidson College. This course can also serve as an introduction to proofs course.

The official textbook for the course was 
\emph{Elementary Number Theory} by James K. Strayer. Some topics that are not covered in these notes were assigned as reading before class. In order to reference these results in the notes, they are provided in \hyperref[other-results]{Appendix A}. The reading assignments are visible by using the \verb|instructornote| option in the \TeX\ file. 

The introduction to proofs used \href{https://danaernst.com/IBL-IntroToProof/pretext/root--1-2.html}{An Introduction to Proof via Inquiry-Based Learning} by Dana C. Ernst, an open source textbook. My best effort has been made to link directly to this resource, although some standard staments and exercies are included in the notes.

Solutions to some problems from Strayer and Ernst are omitted. Solutions to some standard number theory problems from these sources are included.

These notes are also based on my notes teaching the Elementary Number Theory at The Ohio State University. Those courses use \emph{An Introduction to the Theory of Numbers}
by Ivan Niven, Herbert S. Zuckerman, and Hugh L. Montgomery and \emph{Elementary Number Theory} by Gareth A. Jones and J. Mary Jones. These notes were also influenced by \emph{Number Theory: A Lively Introduction with Proofs, Applications, and Stories} by James Pommersheim, Tim Marks, Erica Flapan.

Support for converting these notes to Ximera was provided by a Ximera Flash Grant.
\end{document}