\documentclass{ximera}
\input{../lessonplanheader.tex}

\title{Greatest Common Divisors and Diophantine Equations}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{obj}
	\item Prove the formula for integer solutions to $ax+by=c$.
	\item State when integer solution exist for $a_1x_1+\cdots+a_kx_k=c$.
\end{obj}


\begin{instructorNotes}
	\begin{pre}
		\item[Read] Strayer, Section 6.1
		\item[Turn in] Exercise 2a.
		Find all integer solutions to $18x+28y=10$
	   % \begin{solution}
	   % Notice that $18(-1)+28(1)=10$. This can be done by inspection or the \nameref{euclid-alg}. Then our specific solution is $x_0=-1, y_0=1$ and all solutions have the form \[x=-1+\frac{28n}{2}=-1+14n,\quad y=1-\frac{18n}{2}=1-9n \qquad \textnormal{for all $n\in\Z$.}\]
	   % \end{solution}
	   \end{pre}
\end{instructorNotes}


\begin{lemma}\label{lem:gcd_3case}
	Let $a,b,c\in\Z,$ with $a\neq 0$. Then $(a,b,c)=((a,b),c).$

 	\begin{proof}
 		Let $a,b,c\in\Z,$ with $a\neq 0$. Define $d=(a,b,c)$ and $e=((a,b),c).$ We will show that $d\mid e$ and $e\mid d$. Since the greatest common divisor is positive, we can conclude that $d=e$\footnote{This is not true in general and a common mistake. In general $d=\pm e$}.

 		Since $d=(a,b,c),$ we know $d\mid a, d\mid b,$ and $d\mid c$. By Lemma \ref{lem:gcd_mult}, which we are about to prove, $d\mid (a,b)$. Thus, $d$ is a common divisor of $(a,b)$ and $c$, so $d\mid e$.

 		Since $e=((a,b),c)$, $e\mid  (a,b)$ and $e\mid c$. Since $e\mid (a,b),$ we know $e\mid a$ and $e\mid b$ by Lemma \ref{lem:gcd_mult}. Thus, $e$ is a common divides of $a,b$ and $c$
 	\end{proof}
\end{lemma}


\begin{lemma}\label{lem:gcd_mult}
	Let $a,b\in\Z,$ not both zero. Then any  common divisor of $a$ and $b$ divides the greatest common divisor.

 	\begin{proof}
 		Let $a,b\in\Z,$ not both zero. By \nameref{Bezout}, $(a,b)=am+bn$ for some $n,m\in\Z$. Thus, $d\mid (a,b)$ by linear combination.
 	\end{proof}
\end{lemma}


\begin{lemma}\label{lem:gcd_trans}
 	Let $a,b\in\Z,$ not both zero. Then any divisor of $(a,b)$ is a common divisor of $a$ and $b$.

	\begin{proof}
 		Let $c$ be a divisor of $(a,b)$. Since $(a,b)\mid a$ and $(a,b )\mid b,$ then $c\mid a$ and $c\mid b$ by transitivity.
	\end{proof}
\end{lemma}

%\begin{prop*}
%	Let $a,b,c\in\Z,$ at least one not equal to zero, and $d=(a,b,c)$. Then there exists integers $x,y,z$ such that \[ax+by+cz=d.\]
%\end{prop*}


\begin{proposition}
 	Let $a_1,\dots,a_n\in\Z$ with $a_1\neq 0$.  Then 
		\[
			(a_1,\dots,a_n)
			=((a_1,a_2,a_3,\dots,a_{n-1}),a_n).
		\]
	\begin{proof}
 		Let $k=2$. The since $((a_1,a_2))=(a_1,a_2)$ by the definition of \nameref{defn:gcd} of one integer,  $(a_1,a_2)=((a_1,a_2))$. The $k=3$ case is the first lemma in this section (\ref{lem:gcd_3case}).
 
 		Assume that for all $2\leq k< n$, 
 			\[
				(a_1,\dots,a_k)
				=((a_1,a_2,a_3,\dots,a_{k-1}),a_k).
			\]
		Let $d=(a_1,a_2,a_3,\dots,a_{k}),$
		$e=((a_1,a_2,a_3,\dots,a_{k}),a_{k+1})=(d,a_{k+1}),$ and $f= (a_1,a_2,a_3,\dots,a_{k},a_{k+1}).$ We will show that $e\mid f$ and $f\mid e$. Since both $e$ and $f$ are positive, this will prove that $e=f$.

		Note that $e\mid (a_1,a_2,a_3,\dots,a_{k})$ and $e\mid a_{k+1}$ by definition. 
		Since $(a_1,\dots,a_k)=((a_1,a_2,a_3,\dots,a_{k-1}),a_k)$ by the induction hypothesis, $e\mid(a_1,a_2,a_3,\dots,a_{k-1})$ and $e\mid a_k$ by Lemma \ref{lem:gcd_trans}. Again, by the induction hypothesis, $(a_1,a_2,a_3,\dots,a_{k-1})=((a_1,a_2,a_3,\dots,a_{k-2}),a_{k-1}),$ so $e\mid a_{k-1}$ and $e\mid (a_1,a_2,a_3,\dots,a_{k-2})$ by Lemma \ref{lem:gcd_trans}. Repeat this process until we get $(a_1,a_2,a_3)=((a_1,a_2),a_3)$, so $e\mid a_3$ and $e\mid (a_1,a_2)$ by Lemma \ref{lem:gcd_trans}. Thus $e\mid a_1,a_2,\dots,a_{k+1}$ by repeated applications of Lemma \ref{lem:gcd_trans}. By the generalized version of the Lemma \ref{lem:gcd_mult} on Homework 3, $e\mid f.$

		To show that $f\mid e,$ we note that $f\mid a_1,a_2,\dots, a_k,a_{k+1}$ by definition. Then $f\mid d$ by the generalized version of the Lemma \ref{lem:gcd_mult} on Homework 3. Since $e=(d,a_k),$ we have that $f\mid e$ by Lemma \ref{lem:gcd_mult}.
	\end{proof}
\end{proposition}



\end{document}
