\documentclass{../ximera}
\input{../lessonplanheader.tex}

\title{More algebraic proofs and the Euclidean Algorithm}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{obj}
\item  Understand turning scratch work into proof for algebraic proofs
\item  Approach the floor function problems from Homework 1
\item Prove the Euclidean Algorithm halts and generates the greatest common divisor of two positive integers
\end{obj}

\begin{pre}
 \item[Read] None 
 \item[Turn in] The Learn TeX assignment using a copy of the Homework Template.
\end{pre}
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Example of proof like the floor function (30 minutes)}
%%%%%%%%%%%%%%%%%%%%%%%%%%
I did not go over the entire example in class. This also took the place of in class problems.

Let's explicitly think about the floor function as a function. That is, $f(x):\mathbb{R}\to\mathbb{Z}$ (a function from the real numbers to the integers). Here is a restatement of the homework problem:

\begin{prob}
For each of the following equations, find a domain for $f(x)=\floor{ x }$ make the statement true. Prove your statement. 
	\begin{enumerate}
 		\item $f( x ) + f( x ) =f( 2x)$
		\item $f( x + 3 )  = 3 +f( x)$
		\item $f( x +3 ) = 	3 + x$
	\end{enumerate} 
\end{prob}

Here is a similar problem with proofs where $g:\mathbb{R}\to\mathbb{R}$. Note that the scratch work is one way to think about solving the problem but would not be included in the homework writeup.

\begin{example}
 For each of the following equations, find the full domain for $g(x)=3\sin(\pi x)$ that
 makes the statement true. Prove that the equation is always true on this domain. 
	\begin{enumerate}
 		\item $g( x ) + g( x ) =g( 2x)$
		
		\begin{scratch}
		We want to find a restriction such that $3\sin(\pi x)+3\sin(\pi x)=3\sin(2\pi x)$. Using the double angle formula, we get \[3\sin(2\pi x)=6\sin(\pi x)\cos(2\pi x).\] This means we are actually looking for 
			\begin{align*}
 				6\sin(\pi x)\cos(2\pi x)&=6\sin(\pi x)\\
				\cos(2\pi x)&=1.
			\end{align*}
		\end{scratch}

		\begin{solution}
 		If $x\in\mathbb{Z}$, then $g(x)+g(x)=g(2 x)$.
 
		\begin{proof}
 		Let $x\in\mathbb{Z}$. Then $g(x)+g(x)=3\sin(\pi x)+3\sin(\pi x)=0$ and $g(2x)=3\sin(2\pi x)=0$. Thus, $g(x)+g(x)=g(2 x)$.
		\end{proof}
		\end{solution}

		\item $g( x + 3 )  = 3 +g( x)$
		
		\begin{scratch}
 			We want to find a restriction such that $3\sin(\pi (x+3))=3+3\sin(\pi x)$.
			Using the angle addition formula, we get 
			\[3\sin(\pi x+3\pi))=3\sin(\pi x)\cos(3\pi) +3\cos(\pi x)\sin(3\pi)=-3\sin(\pi x).\] This means we are actually looking for 
			\begin{align*}
 				-3\sin(\pi x)&=3\sin(\pi x)+3\\
				-1&=2\sin(\pi x)\\.
			\end{align*}
		\end{scratch}
		
		
		\begin{solution}
 		If $x=2k+\frac{7}{6}$ or $x=2k-\frac{1}{6}$ for some $k\in\mathbb{Z}$, then $g (x+3)=g(x)+3.$
		
		\begin{proof}
		
		\begin{description} Note that $g(x+3)=3\sin(\pi x+3\pi)=-3\sin(\pi x)$ by the angle addition formula. We will consider two cases:
 			\item{Case 1:} Let $x=2k+\frac{7}{6}$ for some $k\in\mathbb{Z}$. Then 			
			\begin{align*}
 			g(x+3)&=-3\sin(\pi x)\\
			&=-3\sin(2k\pi+\frac{7\pi}{6})\\
			&=\frac{3}{2},
			\end{align*}
			and $g(x)+3=3\sin(2k\pi+\frac{7\pi}{6})+3=\frac{3}{2}$.
			\item{Case 2:} Let $x=2k-\frac{1}{6}$ for some $k\in\mathbb{Z}$. Then 			
			\begin{align*}
 			g(x+3)&=-3\sin(\pi x)\\
			&=-3\sin(2k\pi-\frac{\pi}{6})\\
			&=\frac{3}{2},
			\end{align*}
			and $g(x)+3=3\sin(2k\pi-\frac{\pi}{6})+3=\frac{3}{2}$.
		\end{description}
		Thus, $g(x+3)=g(x)+3$ when $x=2k+\frac{7}{6}$ or $x=2k-\frac{1}{6}$ for some $k\in\Z$.
		\end{proof}
		\end{solution}
		\item Let $h(x)=x^2-1$. For each of the following equations, find the full domain for $h(x)$ that makes the statement true. Prove your statement.
		
		\begin{enumerate}
			\item $h(x+3)=h(x)+3$
			\item $h(x+3)=x+3$
		\end{enumerate}


		\begin{scratch}
			First, $h(x+3)=(x+3)^2-1=x^2+6x+8$.
			
			For (a) 
			\begin{align*}
				x^2+6x+8&= x^2 -1+3\\
				6x& = -6
			\end{align*}
			For (b)
			\begin{align*}
				x^2+6x+8&= x+3\\
				x^2 +5x+5& = 0\\
				x&=\frac{-5\pm\sqrt{5}}{2}
			\end{align*}
		\end{scratch}
		
		\begin{solution}
			
			\begin{enumerate}
				\item For $h(x)=x^2-1,$ $h(x+3)=h(x)+3$ if and only if $x=-1$.
				\begin{proof}
					Let $h(x)=x^2-1$ and $x=-1$. Then $h(x+3)=3=0+3=h(x)+3$.

					To prove that this is the only such $x$, let $h(x+3)=h(x)+3$. Then $x^2+6x+8=x^2+2$, which simplifies to $x=-1$.
				\end{proof}
				\item For $h(x)=x^2-1,$ $h(x+3)=x+3$ if and only if $x=\frac{-5\pm\sqrt{5}}{2}$.
				\begin{proof}
					Let $h(x)=x^2-1$. First we consider the case $x=\frac{-5+\sqrt{5}}{2}$. Then $h(x+3)=\frac{1+\sqrt{5}}{2}=
					\frac{-5+\sqrt{5}}{2}+3=x+3$. 
					
					Next, we consider the case $x=\frac{-5-\sqrt{5}}{2}$. Then $h(x+3)=\frac{1-\sqrt{5}}{2}=
					\frac{-5-\sqrt{5}}{2}+3=x+3$.

					To prove that these are the only such $x$, let $h(x+3)=x+3$. Then $x^2+6x+8=x+3$, which  hasthe solutions  $x=\frac{-5\pm\sqrt{5}}{2}$.
				\end{proof}
			\end{enumerate}
		\end{solution}
	\end{enumerate}
\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Euclidean Algorithm (20 minutes)}

Typically by \emph{Euclidean Algorithm}, we mean  both the algorithm and the theorem that the algorithm always generates the greatest common divisor of two (positive) integers.


\begin{thm*}[Euclidean algorithm]\label{euclid-alg}
	Let $a,b\in\Z$ with $a\geq b>0$. By the \nameref{div-alg}, there exist $q_1,r_1\in\Z$ such that 
	\[a=b q_1+r_1,\quad 0\leq r_1<b.\]
	If $r_1>0$, there exist $q_2,r_2\in\Z$ such that 
	\[b=r_1 q_2+r_2,\quad 0\leq r_2<r_1.\]
	If $r_2>0$, there exist $q_3,r_3\in\Z$ such that 
	\[r_1=r_2 q_3+r_3,\quad 0\leq r_3<r_2.\]
	
	Continuing this process, $r_n=0$ for some $n$. If $n>1$, then $\gcd(a,b)=r_{n-1}$. If $n=1$, then $\gcd(a,b)=b$.
\end{thm*}
\begin{proof}
	 Note that $r_1>r_2>r_3>\dots\geq0$ by construction. If the sequence did not stop, then we would have an infinite, decreasing sequence of positive integers, which is not possible. Thus, $r_n=0$ for some $n$. 
	 
	 When $n=1$, $a=bq+0$ and $\gcd(a,b)=b$.
	 
	 \nameref{lem:gcd-remainders} states that for $a=bq_1+r_1$, $\gcd(a,b)=\gcd(b,r_1)$. This is because any common divisor of $a$ and $b$  is also a divisor of $r_1=a-bq_1$. 
	 
	 If $n>1$, then by repeated application of the \nameref{lem:gcd-remainders}, we have 
	 \[\gcd(a,b)=\gcd(b,r_1)=\gcd(r_1,r_2)=\cdots=\gcd(r_{n-2},r_{n-1})\]
	 Then $r_{n-2}=r_{n-1} q_n+0$. Thus $\gcd(r_{n-2},r_{n-1})=r_{n-1}$.
\end{proof}
	
When using the \nameref{euclid-alg}, it can be tricky to keep track of what is happening. Doing a lot of examples can help.
	
%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%


\end{document}
