<<<<<<< Updated upstream
\documentclass[handout]{ximera}
\input{handoutPreamble}


\date{January 22, 2024}
=======
\documentclass[handout,nooutcomes]{ximera}

\input{../lessonplanheader.tex}
\input{../printstyles.tex}

\classday{January 22}


\handoutTitle
        
\date{\classday, 2024}
>>>>>>> Stashed changes

\begin{document}
\handoutAbstract
\maketitle
<<<<<<< Updated upstream
    \begin{center}%
            {\large \scshape MAT-255-- Number Theory \hfill Spring 2024 \hfill In Class Work January 22}%
        
        {\large
            Your Name: \hrulefill \quad Group Members:\hrulefill \quad \hrulefill
	    \par}%
    \end{center}%
=======
>>>>>>> Stashed changes

\begin{br} 
    Use the division algorithm on $a=47, b=6$ and $a=281, b=13$.
    \begin{solution}
        When $a=47, b=6,$ we have $q=7,$ and $r=5$ since \[47=6(7)+5\quad and \quad 0\leq 5 <6.\]
        When $a=281, b=13,$ we have $q=21,$ and $r=8$ since \[47=13(21)+8\quad and \quad 0\leq 8 <13.\]
    \end{solution}
    \pdfOnly{\ifhandout{
        \vfill}
        \else
        \fi}
\end{br}

\begin{br} Let $a$ and $b$ be nonzero integers. Prove that there exists a unique $q,r\in\Z$ such that 
    \[a=bq+r, \quad 0\leq r <|b|.\]
    \begin{enumerate}
        \item Use the division algorithm to prove this statement as a corollary. That is, use the \emph{conclusion} of the division algorithm as part of the proof.  Use the following outline:
        \begin{enumerate}
            \item  Let $a$ and $b$ be nonzero integers. Since $|b|>0$, the division algorithm says that there exist unique $p,s\in\Z$ such that
            $\inlineAnswer{a=p|b|+s}$
            and
            $\inlineAnswer{0\leq s<|b|}.$
                
            \item There are two cases:
                \begin{enumerate}
                    \item When $\inlineAnswer{b>0},$
                    the conditions are already met, and $r=\inlineAnswer{s}$ and $q=\inlineAnswer{b}.$
                    
                    \item Otherwise,
                    $\inlineAnswer{b<0},$ 
                    $r=\inlineAnswer{s}$ and $q=\inlineAnswer{-b}.$
                \end{enumerate}
            
            \item Since both cases used that the $p,s$ are unique, then $q,r$ are also unique
        \end{enumerate}
        
        \item Use the \emph{proof} of the division algorithm as a template to prove this statement. That is, repeat the steps, adjusting as necessary, but do not use the conclusion.
            \begin{enumerate}
                \item In the proof of the division algorithm, we let $q=\floor{\frac{a}{b}}$. Here we have two cases:
                \begin{enumerate}
                    \item When $\inlineAnswer{b>0},$
                    $q=\inlineAnswer{\floor{\frac{a}{b}}}$ and $r=\inlineAnswer{a-bq}.$
                        
                    \begin{onlineOnly}
                        \begin{hint}
                            The \TeX code for the floor function is \verb|\lfloor ... \rfloor|
                        \end{hint}
                    \end{onlineOnly}
                    as in the proof of the division algorithm. 
  
                    \item When $\inlineAnswer{b<0},$
                    $q=\inlineAnswer{-\floor{\frac{a}{b}}}$ and $r=\inlineAnswer{a-bq}.$
                \end{enumerate}
                
                \item Summarizing these statements, rewrite $q,r$ in terms of $a$ and $b$, as in the original proof of the division algorithm.
            
        
            \item Now use your scratch work and follow the outline of the proof of the division algorithm to provide a new proof \emph{without referencing the division algorithm.}
        
            \pdfOnly{\ifhandout{
                \vfill
                \vfill}
                \else
                \fi}
        \end{enumerate}
    \end{enumerate}
\end{br}

\end{document}








