<<<<<<< Updated upstream
\documentclass[handout]{ximera}
\input{handoutPreamble}


\date{March 27, 2024}
=======
\documentclass[handout,nooutcomes]{ximera}

\input{../lessonplanheader.tex}
\input{../printstyles.tex}

\classday{March 27}


\handoutTitle
        
\date{\classday, 2024}
>>>>>>> Stashed changes

\begin{document}
\handoutAbstract
\maketitle
<<<<<<< Updated upstream
 	\begin{center}%
    	{\large \scshape MAT-255-- Number Theory 
			\hfill Spring 2024 
			\hfill In Class Work March 27}%
    
		{\large Your Name: \hrulefill \quad 
			Group Members:\hrulefill \quad 
			\hrulefill
			\par}%
 	\end{center}%
=======
>>>>>>> Stashed changes
	 
From class March 20:

\begin{tabular}{cll}
	Modulus & Quadratic residues & Quadratic nonresidues\\\hline
	$2$	& $1$ 	& None\\
	$3$	& $1$	& $2$\\
	$5$	& $1,4$	& $2,3$\\
	$7$	& $1,2,4$	& $3,5,6$
\end{tabular}

\begin{proposition}[Proposition 4.5]
	Let $p$ be an odd prime number and $a,b\in\Z$ with $p\nmid a$ and $p\nmid b.$ Then 
	\begin{enumerate}[label=(\alph*)]
		\item $\legendre{a^2}{p}=1$ \label{squares-are-square}
		\item If $a\equiv b\pmod{p}$ then $\legendre{a}{p}=\legendre{b}{p}$ \label{legendre-respects-mod}
		\item $\legendre{ab}{p}=\legendre{a}{p}\legendre{b}{p}$ \label{legendre-mult}
	\end{enumerate}
\end{proposition}

\begin{theorem}[Theorem 4.6]\label{thm:residue-neg1}
	Let $p$ be an odd prime number. Then 
	\[
		\legendre{-1}{p}=
			\begin{cases}
 				1, & p\equiv 1\pmod{4}\\
				-1, & p\equiv 3\pmod{4}
			\end{cases}.
	\]
\end{theorem}

\begin{theorem}[Quadratic reciprocity]\label{quad-rec}
	Let $p$ and $q$ be distinct primes.  
	\begin{enumerate}[label=(\alph*)]
		\item If $p\equiv 1 \pmod{4}$ or $q\equiv 1\pmod{4},$ then $\legendre{p}{q}=\legendre{q}{p}$
 		\item If $p\equiv q \equiv 3 \pmod{4},$ then $\legendre{p}{q}=-\legendre{q}{p}$
	\end{enumerate}
\end{theorem}




\begin{br}[Strayer Chapter 4, Exercise 35]
	Let $p$ be an odd prime number. Prove the following statements the following provided outlines, which will help solve the next problem, as well.
	
	\begin{enumerate}
		\item $\legendre{3}{p}=1$ if and only if $p\equiv \pm1\pmod{12}.$
		\item $\legendre{-3}{p}=1$ if and only if $p\equiv 1\pmod{6}.$	
	\end{enumerate}
\end{br}

\begin{proof}
	\begin{enumerate}
        \item Since $3\equiv \inlineAnswer{3}\pmod{4},$\footnote{In this problem, this step is repetitive, but it is needed when $p\neq3$.} we need two cases for \nameref{quad-rec}. 
            \begin{enumerate}
                \item If $p\equiv 1\pmod{4},$ then $\legendre{3}{p}=\inlineAnswer{\legendre{p}{3}}$ by \nameref{quad-rec}, and $\legendre{p}{3}=1$ if and only if $p\equiv\inlineAnswer{1\pmod{3}}$. Then $p\equiv \inlineAnswer{1}\pmod{12},$ and this is the unique equivalence class modulo $12$ by the Chinese Remainder Theorem.
            
                \item If $p\equiv 3\equiv -1\pmod{4},$ then $\legendre{3}{p}=\inlineAnswer{-\legendre{p}{3}}$ by \nameref{quad-rec}, and $\legendre{p}{3}=-1$ if and only if $p\equiv\inlineAnswer{ 2\equiv-1\pmod{3}}$. Then $p\equiv \inlineAnswer{-1}\pmod{12},$ and this is the unique equivalence class modulo $12$ by the Chinese Remainder Theorem.
            \end{enumerate}

        Therefore, $\legendre{3}{p}=1$ if and only if $p\equiv \pm1\pmod{12}.$ 
            
        \item From Theorem 4.25\ref{legendre-mult}, $\legendre{-3}{p}= \inlineAnswer{\legendre{-1}{p}\legendre{3}{p}}.$
        Again, we have two cases.
            \begin{enumerate}
                \item If $p\equiv 1\pmod{4},$ then $\legendre{-1}{p}=\inlineAnswer{1}$ by \nameref{thm:residue-neg1} and
                $\legendre{3}{p}=\inlineAnswer{\legendre{p}{3}}$ by \nameref{quad-rec}. Thus, $\legendre{-3}{p}=\inlineAnswer{\legendre{p}{3}}=1$ if and only if $p\equiv \inlineAnswer{1\pmod{3}}.$ Then $p\equiv \inlineAnswer{1}\pmod{12},$ and this is the unique equivalence class modulo $12$ by the Chinese Remainder Theorem.
            
                \item If $p\equiv 3\equiv -1\pmod{4},$ then $\legendre{-1}{p}=\inlineAnswer{-1}$ by \nameref{thm:residue-neg1} and
                $\legendre{3}{p}=\inlineAnswer{-\legendre{p}{3}}$ by \nameref{quad-rec}. Thus, $\legendre{-3}{p}=\inlineAnswer{\legendre{p}{3}}=1$ if and only if $p\equiv \inlineAnswer{1\pmod{3}}.$ Then $p\equiv \inlineAnswer{7}\pmod{12},$ and this is the unique equivalence class modulo $12$ by the Chinese Remainder Theorem.
            \end{enumerate}
        Therefore, $\legendre{-3}{p}=1$ if and only if $p\equiv \inlineAnswer{1,7}\pmod{12},$ which is equivalent to $p\equiv 1\pmod{6}.$ 
    \end{enumerate}
\end{proof}

\begin{br}[Strayer Chapter 4, Exercise 36]
	Find congruences characterizing all prime numbers $p$ for which the following integers are quadratic residues modulo $p,$ as done in the previous exercise. 
	
	Outline is provided for the first part.
	\begin{enumerate}
		\item $5$ 
		\item $-5$	
		\item $7$ 
		\item $-7$	
	\end{enumerate}



    \begin{proof}
        \begin{enumerate}
            \item Since $5\equiv\inlineAnswer{1}\pmod{4},$ $\inlineAnswer[1.25 in]{\legendre{5}{p}=\legendre{p}{5}}$ by \nameref{quad-rec}. Then $\legendre{5}{p}=\inlineAnswer[1cm]{\legendre{p}{5}}=1$ if and only if $\inlineAnswer[1.25 in]{p\equiv 1, 4 \pmod{5}}.$

            \pdfOnly{\ifhandout
                \vfill\else\fi}
        \end{enumerate}
    \end{proof}
\end{br}
\end{document}