\documentclass{ximera}
\input{../lessonplanheader.tex}

\title{Nonlinear Diophantine equations}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

\begin{obj}
    \item Define a nonlinear Diophantine equation
\end{obj}


\begin{pre}
    \item[Reading] None
\end{pre}


%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}
	A Diophantine equation is \emph{nonlinear} if it is not linear.
\end{definition}
 
\begin{example}
  
	\begin{enumerate}
 		\item The Diophantine equation $x^2+y^2=z^2$ is our next section. Solutions are called Pythagorean triples.
 		\item Let $n\in\mathbb{Z}$ with $n\geq 3$. The Diophantine equation $x^n+y^n=z^n$ is the subject of the famous Fermat's Last Theorem. We will also prove one case of this.
 		\item Let $n\in\mathbb{Z}$. The Diophantine equation $x^2+y^2=n$ tells us which integers can be represented as the sum of two squares.
 		\item Let $d,n\in\mathbb{Z}$. The Diophantine equation $x^2-dy^2=n$ is known as Pell's equation.
	\end{enumerate}
\end{example}
 
Sometimes we can use congruences to show that a particular nonlinear Diophantine equation has no solutions.
 
\begin{example}
	Prove that $3x^2+2=y^2$ is not solvable.

	\begin{solution}
		Assume that there is a solution. Then any solution to the Diophantine equation is also a solution to the congruence $3x^2+2\equiv y^2 \mod 3$, which implies $2\equiv y^2 \mod 3$, which we know is false. Thus there are no integer solutions to $3x^2+2=y^2$.
	\end{solution}
 \end{example}
 
Note: viewing the same equation modulo $2$ says $x^2\equiv y^2 \mod 2$, which does not give us enough information to prove a solution does not exist--it also is not enough information to conclude a solution exists.
 

\end{document}
