\documentclass{ximera}
\input{../lessonplanheader.tex}

\title{Pythagorean triples}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

\begin{obj}
    \item Define a nonlinear Diophantine equation
    \item Define a primitive Pythagorean triple
	\item Prove the formula for generating primitive Pythagorean triples
\end{obj}

One of the most famous math equations is $x^2+y^2=z^2$, probably because we learn it in high school. We are going to classify all integer solutions to the equation.
 
\begin{definition}
	A triple $(x,y,z)$ of positive integers satisfying the Diophantine equation $x^2+y^2=z^2$ is called \emph{Pythagorean triple}.
\end{definition}
 

\begin{onlineOnly}
	Select the Pythagorean triples:
  
	\begin{selectAll}
	 	\choice[correct] {3,4,5}
	 	\choice[correct]{5,12,13}
	 	\choice{-3,4,5}
	 	\choice[correct]{6,8,10}
	 	\choice{0,1,1}
	\end{selectAll}
\end{onlineOnly}

 
It is actually possible to classify all Pythagorean triples, just like we did for linear Diophantine equations in two variables. To simplify this process, we will work with $x,y,z>0$, and $(x,y,z)=1$. For any given solution of this form, we have that $(-x,y,z),(x,-y,z),(x,y,-z),(-x,-y,z),(x,-y,-z),(-x,y,-z),$ and $(-x,-y,-z)$ are also solutions to the Diophantine equation, as is $(nx,ny,nz)$ for any integer $n$. Thus, we call such a solution a \emph{primitive Pythagorean triple}.  We call $(0,n,\pm n)$ and $(n,0,\pm n)$ the \emph{trival solutions}.
 
\begin{theorem}
For a primitive Pythagorean triple $(x,y,z)$, exactly one of $x$ and $y$ is even.
\end{theorem}
\begin{proof}
 If $x$ and $y$ are both even, then $z$ must also be even, contradicting that $(x,y,z)=1$.
  
 If $x$ and $y$ are both odd, then $z$ is even. Now we can work modulo $4$ to get a contradiction. Since $x$ and $y$ are odd, we have that $x^2\equiv y^2\equiv \answer{1}\pmod 4$. Since $z$ is even, we have that $z^2\equiv \answer{0}\pmod 4$, but $x^2+y^2\equiv \answer{2}\pmod 4$.
  
 Thus, the only remaining option is exactly one of $x$ and $y$ is even.
\end{proof}

\begin{theorem}[Theorem 6.3]\label{thm:form-pyth-trip}
	There are infinitely many primitive Pythagorean triples $x,y,z$ with $y$ even. Furthermore, they are given precisely by the equations
   \begin{align*}
	x&=m^2-n^2\\
	y&=2mn\\
	z&=m^2+n^2
   \end{align*}
   where $m,n\in\mathbb{Z}, m>n>0, (m,n)=1$ and exactly one of $m$ and $n$ is even.
\end{theorem}

\begin{example}

\begin{enumerate}
	\item $m=2$ and $n=1$ satisfy the conditions of $m$ and $n$ in the theorem. This gives $x=\answer{3}, y=\answer{4},z=\answer{5}$.
	\item $m=3$ and $n=2$ gives $x=\answer{5},y=\answer{12}, z=\answer{13}$.
	\item Try with your own values of $m$ and $n$.
   \end{enumerate}
\end{example}
   \begin{proof}
	We first show that given a primitive Pythagorean triple with $y$ even, there exist $m$ and $n$ as described. Since $y$ is  even, $y$ and $z$ are both odd. Moreover, $(x,y)=1,(y,z)=1,$ and $(x,z)=1$. Now,
	\[y^2=z^2-x^2=(x+z)(z-x)\] implies that \[\left(\frac{y}{2}\right)^2=\frac{(x+z)}{2}\frac{(z-x)}{2}.\]
	 
	To show, $\left(\frac{(x+z)}{2},\frac{(z-x)}{2}\right)=1$, let $\left(\frac{(x+z)}{2},\frac{(z-x)}{2}\right)=d.$ Then $d\mid\frac{z+x}{2}$ and $d\mid\frac{z-x}{2}$. Thus, $d\mid\frac{z+x}{2}+\frac{z-x}{2}=z$ and $d\mid\frac{z+x}{2}-\frac{z-x}{2}=x$. Since $(x,z)=1$, we have that $d=1$. Thus, $\frac{(x+z)}{2}$ and $\frac{(z-x)}{2}$ are perfect squares.
	 
	Let \begin{align*}m^2=\frac{(x+z)}{2},\quad n^2=\frac{(z-x)}{2}.\end{align*}
	Then $m>n>0, (m,n)=1, m^2-n^2=x, 2mn=y,$ and $m^2+n^2=z.$ Also, $(m,n)=1$ implies that not both $m$ and $n$ are both even. If both $m$ and $n$  are odd, we have that $z$ and $x$ are both even, but $(x,z)=1$. This proves that every primitive Pythagorean triple has this form.
	 
	Now we prove that given any such $m$ and $n$, we have a primitive Pythagorean triple. First, $(m^2-n^2)^2+
   (2mn)^2=m^4-2m^2n^2+n^4+4m^2n^2=(m^2+n^2)^2.$ We need to show that $(x,y,z)=1$. Let $(x,y,z)=d$. Since exactly one of $m$ and $n$ is even, we have that $x$ and $z$ are both odd. Then $d$ is odd, and thus $d=1$ or $d$ is divisible by some odd prime $p$. Assume that $p\mid d$. Thus, $p\mid x$ and $p\mid z$. Thus, $p\mid z+x$ and $p\mid z-x$. Thus, $p\mid (m^2+n^2)+(m^2-n^2)=2m^2$ and $p\mid (m^2+n^2)-(m^2-n^2)=2n^2.$ Since $p$ is odd, we have that $p\mid m^2$ and $p\mid n^2$, but  $(m,n)=1$, so $d=1$.
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}