\documentclass[letterpaper, 11 pt,handout,hints]{ximera}
\input{../lessonplanheader.tex}

\title{Circle of fifths}


\begin{document}
\begin{abstract}
 Project on frequency ratios for musical cords and continued fractions.
\end{abstract}
\maketitle

Read Strayer Sections 7.2 and 7.3, paying particular attention to the examples. I will use more standard notation than Strayer.

\begin{exploration}
 Give an introduction to continued fraction expansions, with enough detail for classmates to follow the presented problems.

\begin{rubric}
 Present as much as necessarily for classmates to follow: 4 points if individual, 3 points if pair.
\end{rubric}
 
\begin{definition}
 Let $x$ be a positive real number. Then the \emph{continued fraction expansion of $x$} is \[x=a_0+\cfrac{1}{a_1+\cfrac{1}{a_2+\cfrac{1}{a_3+\dots}}}=[a_0;a_1,a_2,\dots]\]
 where $a_1,a_2,\dots$ are positive integers and $a_0$ is a nonnegative integer. 
 
Define the \emph{convergents of $x$} to be the sequence of $\frac{p_n}{q_n}$ where \[\frac{p_n}{q_n}=a_0+\cfrac{1}{a_1+\cfrac{1}{a_2+\cfrac{1}{a_3+\dots+\cfrac{1}{a_n}}}}\]

\begin{problem}
	Use induction to show that $p_{n}=a_n p_{n-1} +p_{n-2}, q_{n}=a_n q_{n-1} +q_{n-2}.$ 
\begin{hint}
 Consider $\cfrac{1}{a_n+\cfrac{1}{a_{n+1}}}$ and use $a_n+\cfrac{1}{a_{n+1}}$ in the induction step.
\end{hint}
\begin{rubric}
4 points if individual, 3 points if pair.
\end{rubric}
\end{problem}

\begin{problem}(If presenting as a pair)
 Chapter 7, Exercise 18
 \begin{rubric}
3 points.
\end{rubric}
\end{problem}
\end{definition}
\end{exploration}

\begin{exploration}
The following problems are from \emph{Number Theory: A Lively Introduction with Proofs, Applications, and Stories} by Erica Flapan, Tim Marks, and James Pommersheim Chapter 14, Continued Fractions, Section 14.3 Approximating Irrational Numbers Using Continued Fractions \cite{theo}.

	Read the scanned notes on Moodle. Present as much as necessarily for classmates to follow.
	
\begin{rubric}
 Present as much as necessarily for classmates to follow: 4 points if individual, 3 points if pair.
\end{rubric}

\begin{problem}
	An acoustically correct major third has frequency ratio $5:4$ 
	\begin{enumerate}
 		\item Show that there do not exist natural numbers $m$ and $n$ such that \[\left(\frac{5}{4}\right)^m=2^n.\]
		Thus, no number of acoustically correct major thirds that can make up a. whole number of octaves. 
		\item Find the first four convergents of $\log_2\left(\frac{5}{4}\right)$
		\item For each of your answer to part (b), find a pair of integers $m$ and $n$ for which $\left(\frac{5}{4}\right)^m=2^n$ is approximately correct.
		\item In the equal-tempered scale, and octave consists of exactly 3 major thirds. To which convergent of $\log_2\left(\frac{5}{4}\right)$ does this correspond?

	\end{enumerate}
\begin{rubric}
Each part: 2 points.
\end{rubric}
\end{problem}
\end{exploration}
\end{document}