\documentclass[letterpaper, 11 pt,handout]{ximera}
\input{../lessonplanheader.tex}

\title{Gaussian Integers}


\begin{document}
\begin{abstract}
 Project on Gaussian Integers.
\end{abstract}
\maketitle

The following problems are from \emph{Number Theory: A Lively Introduction with Proofs, Applications, and Stories} by Erica Flapan, Tim Marks, and James Pommersheim.

	Read the scanned notes on Moodle. Present as much as necessarily for classmates to follow.
	
\begin{rubric}
 	Present as much as necessarily for classmates to follow: 4 points if individual, 3 points if pair.
\end{rubric}

\begin{exploration}
Here I will use slightly more standard notation, which will match iwth the $\Z[\sqrt{d}]$ topic and complex analysis.
 	\begin{definition}
 		The set of \emph{Gaussian integers}, denoted $\Z[i]$ (read ``$\Z$ adjoin $i$") is defined by 
 \[\Z[i]=\{a+bi : a,b\in\Z\},\]
 where $i^2=-1$.
 
 	Let $z=a+bi$ be a Gaussian integer. The \emph{complex  conjugate of $z$} is $\overline{z}=a-bi$. Define the \emph{norm of $z$}, as
 \[N(z)=|z|=z\overline{z}=(a+bi)(a-bi)=a^2+b^2.\]
	\end{definition}

\begin{lem*}[Lemma 13.1.4]\label{norm-mult}
 	Let $r$ and $s$ be Gaussian integers. Then 
 		\[N(rs)=N(r)N(s).\]
 \end{lem*}

\begin{problem}
Prove \nameref{norm-mult}
	\begin{rubric}
 		4 points if individual, 3 points if pair.
	\end{rubric}
\end{problem}

\begin{problem}
Let $d$ and $z$ be Gaussian integers. 
\begin{enumerate}
	\item Prove that if $d\mid z$, then $|d|\mid |z|$.
		\item Prove or provide a counterexample for if $|d|\mid |z|,$ then $d\mid z.$ 
\end{enumerate}
	\begin{rubric}
 		4 points if individual, 3 points if pair.
	\end{rubric}
\end{problem}

\begin{definition}
 Let $p\in\Z[i]$ such that $p$ is not a unit. We say $p$ is \emph{prime} if for every $a,b\in\Z[i],$ $p=ab$ implies that $a$ is a unit or $b$ is a unit.
\end{definition}

\begin{lem*}[Lemma 13.2.6]\label{prime_norm}
 Let $z$ be a Gaussian integer. If $N(z)$ is a prime in the (regular) integers, then $z$ is a prime as a Gaussian integer.
 %
\end{lem*}

\begin{problem}
Prove \nameref{prime_norm}.
	\begin{rubric}
 		4 points if individual, 3 points if pair.
	\end{rubric}
\end{problem}



\begin{definition}
 Let $a,d\in\Z[i]$. We say $a$ and $b$ are \emph{relatively prime} if for all $d\in\Z[i],$ $d\mid a$ and $d\mid b$ implies that $d$ is a unit.
\end{definition}

\begin{thm*}[Division Theorem for Gaussian Integers]\label{div-gauss-int}
 Let $a,d\in\Z[i]$ with $b\neq 0$. Then there exist Gaussian integers $q$ and $r$ such that $a=qb+r$ with $N(r)<N(b)$
 %Exercises 15-17
\end{thm*}

\begin{problem}(If presenting as a pair)
Prove \nameref{div-gauss-int} using Exercises 15-17 in the scanned notes.
	\begin{rubric}
 		5 points.
	\end{rubric}
\end{problem}

\begin{thm*}[Theorem 13.5.4]
 Suppose $p\in\Z$ is prime. Then $p$ is a prime in $\Z[i]$ if and only if $p\equiv 3\pmod{4}$.
 %
\end{thm*}



\begin{thm*}[Theorem 13.5.5]\label{thm:guass-primes}
 Let $z\in\Z[i]$. Then $z$ is a prime Gaussian integer if and only if one of the following conditions holds:
 
\begin{itemize}
 \item $N(z)=2$,
 \item $N(z)$ is a prime integer congruent to $1$ modulo $4$,
 \item $z$ is a unit times a prime integer congruent to $3$ modulo $4$.
\end{itemize}
%
\end{thm*}

\begin{problem}
Prove \nameref{thm:guass-primes}
	\begin{rubric}
 		4 points if individual, 3 points if pair.
	\end{rubric}
\end{problem}
\end{exploration}

\end{document}