\documentclass[letterpaper, 11 pt,handout,hints]{ximera}
\input{../lessonplanheader.tex}

\title{Instructions}\label{finalProjects}
\begin{document}
\begin{abstract} %% This describes your course  
Topics for final presentation in MAT-255 Number Theory, Spring 2024. 
\end{abstract}  
\maketitle
Attendance for each of the presentations is mandatory. 

You may present individually or in pairs. Individual presentations should be 15-20 minutes and pair presentations should be 20-30 minutes. All presentations are on the whiteboard, although you may write on a tablet/slides if you check the technology ahead of time (and be prepared for it to not cooperate). You can have notes with you, and you may rely on them similar to how reliant I am on mine during lectures. But you should prepare this presentation enough that you aren’t always directly copying your notes.

Submit your ranking of choices by the start of class on April 24. If you choose to work with a partner, only one person has to submit the rankings. I will send out the project assignments after class on Wednesday.

I encourage you to ask for help leading up to the presentation practice your presentation for time and accuracy, including presenting on the board.

Rubrics: 
\begin{description}
 	\item[Content (20 points)] Accuracy of proof, correct usage of terms, appropriate for the audience and within the scope of
this class. The number of points per problem is on the individual topic assignments.
	\item[Organization (10 points)] Overall information is presented in a logical sequence, including any necessary background information or examples without extraneous information.
	\item[Presentation (5 points)] Well-prepared and well-rehearsed, spoken loud enough to hear easily, spoken and not read, good boardwork
	\item[Pacing (5 points)] Slowing down and being more detailed in tougher to understand portions of the presentation, following the time guidelines, if presenting in a pair splitting time and content equally with your partner
	\item[Overall topic fluency (5 points)] Able to answer 1-2 questions pertaining to presentation or clarification questions from classmates
	\item[Peer-feedback (5 points)] Attends all presentations and fills out feedback and question forms
\end{description}

\end{document}