\documentclass[letterpaper, 11 pt,handout]{ximera}
\input{../lessonplanheader.tex}

\title{Other number systems: $\Z[\sqrt{d}]$}


\begin{document}
\begin{abstract}
 Project on $\Z[\sqrt{d}].$
\end{abstract}
\maketitle

\begin{definition}
 Let $d\in\Z$ such that $d$ is not a perfect square. Define $\Z[\sqrt{d}]=\{a+b\sqrt{d}:a,b\in\Z\},$ pronounced ``$Z$ adjoin square-root $d$" If $d<0,$ then $\Z[\sqrt{d}]$ is a set of complex numbers. We will use the regular definition of addition and multiplication.
\end{definition}

\begin{definition}
	 Let $d\in\Z$ such that $d$ is not a perfect square, and let $z\in\Z[\sqrt{d}]$ where $z=a+b\sqrt{d}.$ The \emph{conjugate of $z$} is $\overline{z}=a-b\sqrt{d},$ and the \emph{norm of $z$} is $|z|=z\overline{z}=a^2-db^2.$
	 
	 Strayer uses the notation $z^\prime$ for the conjugate.
\end{definition}

\begin{exploration}
\begin{problem}
 	Strayer Chapter 1, Exercise 77 explores the case $d=-10.$ Then $\Z[\sqrt{d}]$ is not have unique factorization. Do this problem. 
\begin{rubric}
 Parts (a) and (b): 3 points each if individual, 2 points each if pair. Part (c): 2 points.
 
 Problem total: 8 points if individual, 6 points if pair.
\end{rubric}
\end{problem}

\begin{problem}
	If $x,y\in\Z[\sqrt{d}],$ prove that $|xy|=|x||y|.$ If $\frac{x}{y}\in\Z[\sqrt{d}],$ prove that $\left|\frac{x}{y}\right|=\frac{|x|}{|y|}.$ 
\begin{rubric}
2 points.
\end{rubric}
\end{problem}

\begin{problem}(If presenting as a pair) 
	\begin{enumerate}
 		\item Prove that the norm of any nonzero element (ie, $a\neq 0$ or $b\neq 0$ of $\Z[\sqrt{-5}]$ is positive.
		
		\item When, if ever, does $|a+b\sqrt{-5}|=1$?
		
		\item When, if ever, does $|a+b\sqrt{-5}|=13$?
	\end{enumerate}
\begin{rubric}
4 points.
\end{rubric}	
\end{problem}



\begin{problem}
	Strayer Chapter 8, Student Project 6.
\begin{rubric}
 Part (a): 2 points. Parts (b) and (c): 4 points each if individual, 3 points each if pair.
 
 Problem total: 10 points if individual, 8 points if pair.
\end{rubric}
\end{problem}
\end{exploration}



\end{document}