\documentclass{../ximera}
\input{../lessonplanheader.tex}

\title{Arithmetic progressions and introduction to Congruences}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{obj}
    \item State and prove facts about prime factorizations using the Fundamental Theorem of Arithmetic
    \item Prove there are infinitely many primes of the form $4n+3$.
    \item Prove a given set is an equivalence relation
\end{obj}

\begin{pre}
    \item[Reading] Strayer, Appendix B
    \item[Turn in] Let $R$ be the equivalence relation on $\R$
    defined by
    \[
   [a]=\{b\in\R:\sin(a)=\sin(b) \textnormal{ and }\cos(a)=\cos(b)\}.\]
     Prove that $R$ is an equivalence relation on $\R.$
   Describe the equivalence classes on $\R$
   
   \begin{solution}
        Since $\sin(a)=\sin(a)$ and $\cos(a)=\cos(a),$ the relation $R$ is reflexive.

        If $\sin(a)=\sin(b)$ and $\cos(a)=\cos(b),$ the $\sin(b)=\sin(a)$ and $\cos(b)=\cos(a),$ so the relation is symmetric.

        If $\sin(a)=\sin(b)$ and $\cos(a)=\cos(b),$ $\sin(b)=\sin(c)$ and $\cos(b)=\cos(c),$ then $\sin(a)=\sin(c)$ and $\cos(a)=\cos(c)$ is transitive.

        Note that $\sin(a)=\sin(b)$ if $b=a+2\pi k$ or $b=-a+\pi+2\pi k$ for some $k\in\Z,$ and  $\cos(a)=\cos(b)$ if $b=a+2\pi k$ or $b=-a+2\pi k$ for some $k\in\Z.$ These conditions are both true with $b=a+2\pi k$. Thus, for $a\in[0,2\pi),$
        \[
            [a]=\{\dots,a-4\pi,a-2\pi,a,a+2\pi,a+4\pi,\dots\}.
        \]
       \end{solution}
\end{pre}
%%%%%%%%%%%%%%%%%%%%%%%%%


% \begin{br}[Strayer Chapter 1, Exercise 36]
%     \begin{enumerate}[label=(\alph*)] %enumerate creates a numbered list, label=(\alph*) says to use lower case numbers in parentheses
%         \item Do there exist integers $x$ and $y$ such that $x + y = 100$ and $(x, y) = 8$?
        
        
%         \begin{solution}
%             On Homework 3.
%         \end{solution}
 
%         \item Prove that there exist infinitely many pairs of integers $x$ and $y$ such that $x + y = 87$ and $(x, y) = 3$.
 
%         \begin{scratch}
%          Note that $87=\answer{3(29)}$.
%          To ensure that $(x,y)=3,$ not just $3\mid x$ and $3\mid y,$ let $x=3n$ where $\answer{29}\nmid n$.
        
%         \end{scratch}
 
%         \begin{proof}
         
%          Let $x\in\Z$ with $\answer{\textnormal{On Homework 3}}$. Let $y= \answer{}.$ Then $3\mid y$ by  $\answer{\textnormal{On Homework 3}}$. Then $(x,y)=3$ since $\answer{\textnormal{On Homework 3}}$. Thus, there are infinitely many $x,y\in\Z$ $\answer{\textnormal{On Homework 3}}$.
         
%         \end{proof}
    
%     \end{enumerate}
% \end{br}



% \begin{br}[Strayer Chapter 1, Exercise 38] Let $a$ and $b$ be relatively prime integers. Prove that $(a+b,a-b)$ is either $1$ or $2$.
	

%     Using the hint in the back of the book and Exercise 37, which states $(ca,cb)=|c|(a,b)$ for all $a,b\in\Z,$ not both $0.$ 
% \begin{proof}
% 	Let $(a+b,a-b)=d$ and note that $d\mid (a+b)+(a-b)$ and $d\mid (a+b)-(a-b)$ by $\answer{\textnormal{linear combination}}$
%     $\answer{\textnormal{On Homework 3}}$
% \end{proof}
% \end{br}


%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Prime factorizations (20 minutes)}
%%%%%%%%%%%%%%%%%%%%%%%%%%
Note on $m^4-n^4=(m^2-n^2)(m^2+n^2)$: In order to sho w this is not prime, must prove that the factors cannot be $1$ and the number itself. Hint: show that if one of the factors is $1$ the other is $1$ or $0$ (or $-1$).


\begin{cor*}[Corollary 1.20]\label{cor:lcm-gcd}
 Let $a,b\in\Z$ with $a,b>0$. Then $[a,b]=ab$ if and only if $(a,b)=1$.
\end{cor*}
A note on ``if and only if" proofs: 
\begin{itemize}
 \item You can do two directions: 
\begin{itemize}
 \item  If $[a,b]=ab$, then $(a,b)=1$.
 \item  If $(a,b)=1$, then $[a,b]=ab$.
\end{itemize}
\item Sometimes you can string together a series of ``if and only if statements." Definitions are always ``if and only if," even though rarely stated that way. For example, an integer $n$ is even if and only if there exist an integer $m$ such that $n=2m$:
\begin{itemize}
 \item An integer $n$ is even if and only if $2\mid n$ (definition of even) 
 \item if and only if there exist an integer $m$ such that $n=2m$ (definition of $2\mid n$).
\end{itemize}
\end{itemize}

\begin{thm*}[Dirichlet's Theorem]\label{thm:dirichlet}
 Let $a,b\in\Z$ with $a,b>0$ and $(a,b)=1$. Then the arithmetic progression \[a,a+b, a+2b, \dots, a+nb,\dots\]
 contains infinitely many primes.
\end{thm*}
Surprisingly, this proof involves complex analysis. The statement that there are infinitely many prime numbers is the case $a=b=1$.


\begin{warning}
    You may not use this result to prove special cases, ie, specific values of $a$ and $b$.
\end{warning}


\begin{lem*}[Lemma 1.23]\label{lem:prod-goodprimes}
 If $a,b\in\Z$ such that $a=4m+1$ and $b=4n+1$ for some integers $m$ and $n$, then $ab$ can also be written in that form.
\end{lem*}
We will not go over the proof in class.
\begin{proof}
 Let $a=4m+1$ and $b=4n+1$ for some integers $m$ and $n$. Then 
\begin{align*}
ab&=(4m+1)(4n+1)\\
&=16mn+4m+4n+1\\
&=4(4mn+m+n)+1.\qedhere
\end{align*}
\end{proof}

\begin{prop*}[Proposition 1.22]\label{prop:inf-badprimes}
 There are infinitely may prime numbers expressible in the form $4n+3$ where $n$ is a nonnegative integer.
\end{prop*}
\begin{proof}
 (Similar to the proof that there are infinitely many prime numbers). Assume, by way of contradiction, that there are only finitely many prime numbers of the form $4n+3$, say $p_0=3, p_1,p_2,\dots, p_r$, where the $p_i$ are distinct. Let $N=4p_1 p_2 \cdots p_r+3$. If every prime factor of $N$ has the form $4n+1$, then so does $N$, by repeated applications of \nameref{lem:prod-goodprimes}. Thus, one of the prime factors of $N$, say $p$, have the for $4n+3$. We consider two cases:
 
\begin{description}
 \item[Case 1, $p=3$:] If $p=3$, then $p\mid N-3$ by linear combinations. Then $p\mid 4 p_1p_2\cdots p_r$. Then by \nameref{cor:irreducible-prime}, either $3\mid 4$ or $3\mid p_1p_2\cdots p_r$. This implies that $p\mid p_i$ for some $i=1,2,\dots,r$. However, $p_1,p_2,\dots,p_r$ are distinct primes not equal to $3$, so this is not possible. Therefore, $p\neq 3$.
 
 \item[Case 2, $p=p_i$ for some $i=1,2,\dots, r$:] If $p=p_i$, then $p\mid N-4p_1p_2\cdots p_r$ by linear combinations. Then $p\mid 3$. However, $p_1,p_2,\dots,p_r$ are distinct primes not equal to $3$, so this is not possible. Therefore, $p\neq p_i$ for $i=1,2,\dots,r$.
\end{description}
Therefore, $N$ has a prime divisor of the form $4n+3$ which is not on the list $p_0,p_1,\dots,p_r$, which contradicts the assumption that $p_0,p_1,\dots,p_r$ are all primes of this form. Thus, there are infinitely many primes of the form $4n+3$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Equivalence Relation Practice (10 minutes)}
%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{br}
    Prove that \[[a]=\{b\in\Z: 3\mid(a-b)\}\] is an equivalence relation on $\Z.$
  \end{br}
  
  \begin{proof}
    Let $a,b\in\Z$. We must show that the relation is reflexive, symmetric, and transitive.
  
     To show the relation is reflexive, we must show $a\in \{b\in\Z: 3\mid(a-b)\}.$ Since $\answer{3\mid a-a=0}$,\\
     $a\in \{b\in\Z: 3\mid(a-b)\}.$
  
     To show the relation is symmetric, we must show that if $x\in \{b\in\Z: 3\mid(a-b)\},$ then $a\in \{b\in\Z: 3\mid(x-b)\}.$ If $x\in \{b\in\Z: 3\mid(a-b)\},$ then $\answer{\textnormal{there exists $k\in\Z$ such that $3k=a-x$}}$. Therefore, $\answer{-3k=b-a}$
     and $a\in \{b\in\Z: 3\mid(x-b)\}.$
  
     To show the relation is transitive, we must show that if $x\in \{b\in\Z: 3\mid(a-b)\}$ and $y\in \{b\in\Z: 3\mid(x-b)\},$ then $x\in \{b\in\Z: 3\mid(a-b)\}.$ If $x\in \{b\in\Z: 3\mid(a-b)\},$ then $\answer{\textnormal{there exists $k\in\Z$ such that $3k=a-x$}}$. 
     Similarly, if $y\in \{b\in\Z: 3\mid(x-b)\},$ then $\answer{\textnormal{there exists $m\in\Z$ such that $3m=x-y$}}$.
     Therefore, $\answer{3(m+k)=a-k}$
     and $y\in \{b\in\Z: 3\mid(a-b)\}.$ $\answer{\textnormal{Since the relation is reflexive, symmetric, and transitive, it is an equivalence relation.}}$
  \end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%


\end{document}
