\documentclass{ximera}
\input{../lessonplanheader.tex}

\title{Prime factorizations}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

\begin{obj}
    \item State and prove facts about prime factorizations using the Fundamental Theorem of Arithmetic
    \item Prove there are infinitely many primes of the form $4n+3$.
\end{obj}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% Note on $m^4-n^4=(m^2-n^2)(m^2+n^2)$: In order to sho w this is not prime, must prove that the factors cannot be $1$ and the number itself. Hint: show that if one of the factors is $1$ the other is $1$ or $0$ (or $-1$).


\begin{corollary}\label{cor:lcm-gcd}
 Let $a,b\in\Z$ with $a,b>0$. Then $[a,b]=ab$ if and only if $(a,b)=1$.
\end{corollary}
A note on ``if and only if" proofs: 
\begin{itemize}
 \item You can do two directions: 
\begin{itemize}
 \item  If $[a,b]=ab$, then $(a,b)=1$.
 \item  If $(a,b)=1$, then $[a,b]=ab$.
\end{itemize}
\item Sometimes you can string together a series of ``if and only if statements." Definitions are always ``if and only if," even though rarely stated that way. For example, an integer $n$ is even if and only if there exist an integer $m$ such that $n=2m$:
\begin{itemize}
 \item An integer $n$ is even if and only if $2\mid n$ (definition of even) 
 \item if and only if there exist an integer $m$ such that $n=2m$ (definition of $2\mid n$).
\end{itemize}
\end{itemize}

\begin{thm*}[Dirichlet's Theorem]\label{thm:dirichlet}
    Let $a,b\in\Z$ with $a,b>0$ and $(a,b)=1$. Then the arithmetic progression \[a,a+b, a+2b, \dots, a+nb,\dots\]
    contains infinitely many primes.
\end{thm*}


\begin{remark}
    Surprisingly, this proof involves complex analysis. The statement that there are infinitely many prime numbers is the case $a=b=1$.
\end{remark}

\begin{warning}
    You may not use this result to prove special cases, ie, specific values of $a$ and $b$.
\end{warning}


\begin{lemma}\label{lem:prod-goodprimes}
 If $a,b\in\Z$ such that $a=4m+1$ and $b=4n+1$ for some integers $m$ and $n$, then $ab$ can also be written in that form.
\begin{proof}
 Let $a=4m+1$ and $b=4n+1$ for some integers $m$ and $n$. Then 
\begin{align*}
ab&=(4m+1)(4n+1)\\
&=16mn+4m+4n+1\\
&=4(4mn+m+n)+1.\qedhere
\end{align*}
\end{proof}
\end{lemma}

\begin{prop*}[Proposition 1.22]\label{prop:inf-badprimes}
 There are infinitely may prime numbers expressible in the form $4n+3$ where $n$ is a nonnegative integer.
\end{prop*}
\begin{proof}
 (Similar to the proof that there are infinitely many prime numbers). Assume, by way of contradiction, that there are only finitely many prime numbers of the form $4n+3$, say $p_0=3, p_1,p_2,\dots, p_r$, where the $p_i$ are distinct. Let $N=4p_1 p_2 \cdots p_r+3$. If every prime factor of $N$ has the form $4n+1$, then so does $N$, by repeated applications of \nameref{lem:prod-goodprimes}. Thus, one of the prime factors of $N$, say $p$, have the for $4n+3$. We consider two cases:
 
\begin{description}
 \item[Case 1, $p=3$:] If $p=3$, then $p\mid N-3$ by linear combinations. Then $p\mid 4 p_1p_2\cdots p_r$. Then by \nameref{cor:irreducible-prime}, either $3\mid 4$ or $3\mid p_1p_2\cdots p_r$. This implies that $p\mid p_i$ for some $i=1,2,\dots,r$. However, $p_1,p_2,\dots,p_r$ are distinct primes not equal to $3$, so this is not possible. Therefore, $p\neq 3$.
 
 \item[Case 2, $p=p_i$ for some $i=1,2,\dots, r$:] If $p=p_i$, then $p\mid N-4p_1p_2\cdots p_r$ by linear combinations. Then $p\mid 3$. However, $p_1,p_2,\dots,p_r$ are distinct primes not equal to $3$, so this is not possible. Therefore, $p\neq p_i$ for $i=1,2,\dots,r$.
\end{description}
Therefore, $N$ has a prime divisor of the form $4n+3$ which is not on the list $p_0,p_1,\dots,p_r$, which contradicts the assumption that $p_0,p_1,\dots,p_r$ are all primes of this form. Thus, there are infinitely many primes of the form $4n+3$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%


\end{document}
