\documentclass{ximera}
\input{../lessonplanheader.tex}

\title{Sums and Differences of Squares}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{pre}
    \item[Reading and Turn in:] \href{https://ximera.osu.edu/elementarynt/MATH4573Online/ElementaryNumberTheory/MATH4573Online/April10/April10}{This fill-in-the-blank version} of the proof of Sum of Three Squares
\end{pre}


\begin{br}[Chapter 6, Exercise 34]
	Prove that a positive integer can be written as the difference of two squares of integers if and only if it is not of the form $4n + 2$ for some $n\in\Z.$
	\begin{solution}
	\begin{description}
	 	\item[($\Rightarrow$)] We will show that if a positive integer can be written as the difference of two squares of integers, then it is not of the form $4n + 2$ for some $n\in\Z.$

	 	(Proof)
	  
		\item[($\Leftarrow$)] We will show that any positive integer not of the form $4n + 2$ for some $n\in\Z$ can be written as the difference of two squares of integers. 
	  
		First, we will show that if $a$ and $b$ are positive integers that can be written as the difference of two squares of integers, then so can $ab.$
		
		(Proof)
	

		Now we will show that every odd prime can be written as the difference of two squares of integers. Let $p$ be an odd prime. Then $p=x^2-y^2=(x-y)(x+y)$ when $x=\answer{\frac{p+1}{2}}$ and $y=\answer{\frac{p-1}{2}}.$
		Therefore every odd number can be written as the difference of two squares since \pdfOnly{\rule{2 cm}{.4pt}.} %\onlineOnly{Think about why this is true.}
	
		It remains to show that every positive integer of the form $4n$ for some $n\in\Z$ can be written as the difference of two squares of integers. Why is this the only remaining case?
	
		Similar to the odd prime case, $4n=x^2-y^2=(x-y)(x+y)$ when $x=\answer{n+1}$ and $y=\answer{n-1}$ or when $x=\answer{\frac{n}{2}-2}$ and $y=\answer{\frac{n}{2}+2}$ (solutions are not unique).
	\end{description}
\end{solution}
\end{br}

If there is time

\begin{br}[Chapter 6, Exercise 14d]
	Let $x,y,z$ be a primitive Pythagorean triple with $y$ even. Prove that $x+y\equiv x-y\equiv 1,7\pmod{8}.$

	\begin{hint}
		First show that $x+y\equiv x-y\pmod{8}.$
	\end{hint}


	\begin{solution}
		Let $x,y,z$ be a primitive Pythagorean triple with $y$ even. 
		Then from \nameref{thm:form-pyth-trip}, there exist positive integers $m,n$ with $m>n, (m,n)=1$ and exactly one of $m,n$ even such that 
		\begin{align*}
			x&=m^2-n^2\\
			y&=2mn\\
			z&=m^2+n^2.
		\end{align*}
		Thus, $4\mid y$ and $y\equiv -y\pmod{8}.$ Adding $x$ to both sides of the congruence gives $x+y\equiv x-y\pmod{8}.$

		Since exactly one of $m,n$ even, $x\equiv 0-1, 4-1, 1-0, 1-4\pmod{8}.$ When $x\equiv \pm 1\pmod{8},$ $4\mid mn$ and thus $y=2mn\equiv 0\pmod{8}.$ Thus,
		$x+y\equiv x-y\equiv \pm 1\pmod{8}.$ 
		When $x\equiv \pm 3\pmod{8},$ $4\nmid mn$ and thus $y=2mn\equiv 4\pmod{8}.$ Thus,
		$x+y\equiv x-y\equiv \pm 1\pmod{8}.$ 
	\end{solution}
\end{br}



\end{document}
