\documentclass{../ximera}
\input{../lessonplanheader.tex}

\title{Induction}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{obj}
\item Construct a proof by induction
\end{obj}

The following reading assignment covers the basics on proof by induction:

  \begin{pre}
    \item[Read] Strayer Appendix A.1: The First Principle of Mathematical Induction  or Ernst \href{https://danaernst.com/IBL-IntroToProof/pretext/sec_Intro_to_Induction.html}{Section 4.1} and \href{https://danaernst.com/IBL-IntroToProof/pretext/sec_More_on_Induction.html}{Section 4.2}
   
    \item[Turn in] Strayer Exercise Set A, Exercise 1a. If $n$ is a positive integer, then 
      \[1^2+2^2+3^2+\cdots+n^2=\frac{n(n+1)(2n+1)}{6}.\]
  
  \end{pre}


\begin{br}
Theorems in Ernst \href{https://danaernst.com/IBL-IntroToProof/pretext/sec_Intro_to_Induction.html}{Section 4.1} 
 

\begin{thm*}[Ernst Theorem 4.5]
For all $n\in\mathbb{N}$, 3 divides $4^{n}-1$.
\end{thm*}
\begin{solution}
We proceed by induction.  When $n=1,$ $3\mid 4^n-1=3$. Thus, the statement is true for $n=1.$

Now assume $k\geq 1$ and the desired statement is true for $n=k$. Then the induction hypothesis is \[3\mid 4^k-1.\]
By the definition of \nameref{defn:divides}, there exists $m\in\Z$ such that $3m=4^k-1.$ In other words, $3m+1=4^k$. Multiplying both sides by $4$ gives $12m+4=4^{k+1}$. Rewriting this equation gives $3(4m+1)=4^{k+1}-1$. Thus, $3\mid 4^{k+1}-1$, and the desired statement is true for $n=k+1$. By the (first) principle of mathematical induction, the statement is true for all positive integers, and the proof is complete.
\end{solution}

 \begin{thm*}[Ernst Theorem 4.7]
 Let $p_{1}, p_{2}, \ldots, p_{n}$ be $n$ distinct points arranged on a circle.  Then the number of line segments joining all pairs of points is $\frac{n^{2}-n}{2}$.
 \end{thm*}
\begin{solution}
 We proceed by induction. When $n=1$, there is only one point, so there are no lines connecting pairs of points. Additionally, $\frac{1^2-1}{2}=0$.\footnote{Alternately, you could use $n=2$ for the base case. Then there is one line connecting the only pair of points and $\frac{2^2-2}{2}=1$}
 
 Now assume $k\geq 1$ and the desired statement is true for $n=k$. Then the induction hypothesis is for $k$ distinct points arranged in a circle, the number of line segments joining all pairs of points is $\frac{k^{2}-k}{2}$. Adding a $(k+1)^{st}$ point on the circle will add an additional $k$ line segments joining pairs of points, one for each existing point. Note that 
 \[ 
 	\frac{k^{2}-k}{2}+k=\frac{k^{2}+k}{2}=
	\frac{k^2+k+k+1-(k+1)}{2}=\frac{(k+1)^{2}-(k+1)}{2}\qedhere
 \]
\end{solution}
\end{br}

\begin{br}
[Strayer Exercise 1]. Use the first principle of mathematical induction to prove each statement.
  \begin{enumerate}
    \item If $n$ is a positive integer, then 
  \[1^3+2^3+3^3+\cdots+n^3=\frac{n^2(n+1)^2}{4}.\]
  
    \item If $n$ is an integer with $n\geq 5,$ then \[2^n>n^2.\]
 
  \end{enumerate}
\end{br}







%\begin{br} Let $a_1,a_2,\dots,a_n\in\Z$ with $a_1\neq 0$. Prove that \[\gcd(a_1,\dots,a_n)=\gcd(\gcd(a_1,a_2,a_3,\dots,a_{n-1}),a_n).\]
%\end{br}



\end{document}
