\documentclass{../ximera}
\input{../lessonplanheader.tex}

\title{Primes}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{obj}
\item  Every integer greater than 1 has a prime divisor.
\item  Prove that there are infinitely many prime numbers
\end{obj}

\begin{pre}
 \item[Read] Strayer, Section 1.2
 \item[Turn in] 
\begin{itemize}
 \item The proof method for Euclid's infinitude of primes is an important method. Summarize this method in your own words.
 
\begin{solution}
 Summaries will vary
\end{solution}
 \item Identify any other new proof methods in this section
 
\begin{solution}
 Proof by construction may be new to some students. Students also identified: 
\begin{itemize}
 \item Introducing a variable to aid in proof
 \item Without loss of generality
 
\end{itemize}
\end{solution}
 \item Exercise 22. Prove that 2 is the only even prime number.
 
\begin{solution}
 Assume that there exists another even prime number, call it $p$. Then there exists $2\mid p$ by the definition of even, but that implies that $p=2$ by the definition of prime. Thus, $2$ is the only even prime number.
\end{solution}
\end{itemize}
\end{pre}
%%%%%%%%%%%%%%%%%%%%%%%%%



\begin{defn}[prime and composite]
An integer $p>1$ is \emph{prime} if the only positive divisors of $p$ are $1$ and itself. An integer $n$ which is not prime is \emph{composite}. 
\end{defn}

Why is $1$ not prime?

%It has to do with units and invertibility. The number $1$ holds a special place. It is the multiplicative identity, i.e., anything multiplied by $1$ is just that thing again. Something is said to be invertible in a ``group" (more on that later) if there exist something, which, when multiplied to it, gives you $1$. How many invertible elements are there among the integers? Just two. $1$ and $-1$. And that's the key. If we want to extend our results from positive integers to non-zero integers, we often just need to take into account $\pm 1$. That sounds obvious, but it turns out to be surprisingly critical and yet non-intuitive when we start moving from real integers to complex ones.

\begin{lem*}[Lemma 1.5]\label{lem:prime-composite}
 Every integer greater than 1 has a prime divisor.
\end{lem*}

We will not go over this proof in class.

\begin{proof} 
 Assume by contradiction that there exists $n\in\Z$ greater than 1 with no prime divisor. By the \nameref{well-order}, we may assume $n$ is the least such integer. By definition, $n\mid n$, so $n$ is not prime. Thus, $n$ is composite and there exists $a,b\in\Z$ such that $n=ab$ and $1<a<n$, $1<b<n$. Since $a<n$, then it has a prime divisor $p$. But since $p\mid a$ and $p\mid n$, $p\mid n$. This contradicts our assumption, so no such integer exists.
\end{proof}


\begin{thm*}[Euclid's Infinitude of Primes]\label{thm:inf-primes}(Theorem 1.6)
 There are infinitely many prime numbers.
\end{thm*}
\begin{proof}
 Assume by way of contradiction, that there are only finitely many prime numbers, so $p_1,p_2,\dots,p_n$. Consider the number $N=p_1p_2\cdots p_n +1$. Now $N$ has a prime divisor, say, $p$, by \nameref{lem:prime-composite}. So $p=p_i$ for some $i$, $i=1,2,\dots,n$. Then $p\mid N-p_1p_2\dots p_n$, which implies that $p\mid 1$, a contradiction. Hence, there are infinitely many prime numbers.
\end{proof}

Another important fact is there are arbitrarily large sequences of composite numbers. Put another way, there are arbitrarily large gaps in the primes. Another important proof method, which is a \emph{constructive proof}:

\begin{prop*}[Proposition 1.8]\label{prop:gaps-primes}
 For any positive integer $n$, there are at least $n$ consecutive positive integers.
\end{prop*}
\begin{proof}
 Given the positive integer $n$, consider the $n$ consecutive positive integers \[(n+1)!+2, (n+1)!+3,\dots, (n+1)!+n+1.\]
 Let $i$ be a positive integer such that $2\leq i\leq n+1$. Since $i\mid (n+1)!$ and $i\mid i$, we have \[i\mid(n+1)! +i,\quad 2\leq i\leq n+1\] by linear combination (\nameref{lem:linear-combo}). So each of the $n$ consecutive positive integers is composite.
\end{proof}

  \begin{br} Let $n$ be a positive integer with $n\neq 1$. Prove that if $n^2+1$ is prime, then $n^2+1$ can be written in the form $4k+1$ with $k\in\Z$.
  \begin{solution}
   Assume that $n$ is a positive integer, $n\neq 1,$ and $n^2+1$ is prime. If $n$ is odd, then $n^2$ is odd, which would imply $n^2+1=2,$ the only even prime. However, $n\neq 1$ by assumption. Thus, $n$ is even. 

    By definition of even, there exists $j\in\Z$ such that $n=2k$ and $n^2=4j^2$. Thus, $n^2+1=4k+1$ when $k=j^2.$
  \end{solution}
\end{br}

  \begin{br}
  Prove or disprove the following conjecture, which is similar to Conjecture 1:\\
      \textbf{Conjecture:} There are infinitely many prime number $p$ for which $p+2$ and $p+4$ are also prime numbers.

  
  \begin{solution}
    On Homework 2.
  \end{solution}
    \end{br}



%%%%%%%%%%%%%%%%%%%%%%%%%%


\end{document}
